---
title: "Read in data, look at basic statistics, generate Table 1"
author: "Boca et al"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load all libraries and functions

```{r}
library(knitr)
library(RCurl) ##for getURL

source("functions.R")
```

## Load data, add age categories

Read in data from csv file on Github:

```{r}
file_url <- getURL("https://raw.githubusercontent.com/SiminaB/DMD-metabolomics/master/S1_Table.csv")
MasterFrame <- read.csv(text=file_url)
```

Take a look at the dataset:

```{r}
dim(MasterFrame)
head(colnames(MasterFrame))
tail(colnames(MasterFrame))

MasterFrame[1:5,1:5]
```

Create vector of metabolites and get number of metabolites:

```{r}
##metabolite IDs all have the format (from XCMS) of M.mass.T.retentiontime.n/p 
##(for negative or positive mode) 
mets <- colnames(MasterFrame)[grep("M[[:digit:]]+T[[:digit:]]+",colnames(MasterFrame))]
head(mets)
tail(mets)
length(mets)
```

Create age categories:

```{r}
range(MasterFrame$Age)

MasterFrame$Category <- ""

MasterFrame$Category[MasterFrame$Age >= 4 & MasterFrame$Age < 7] <- 
  "4-7 years"
MasterFrame$Category[MasterFrame$Age >= 7 & MasterFrame$Age < 11] <- 
  ">7-11 years"
MasterFrame$Category[MasterFrame$Age >= 11 & MasterFrame$Age < 18] <- 
  ">11-18 years"
MasterFrame$Category[MasterFrame$Age >= 18 & MasterFrame$Age < 29] <- 
  ">18-29 years"

MasterFrame$Category <- 
  factor(MasterFrame$Category, 
         levels = c("4-7 years", 
                    ">7-11 years", 
                    ">11-18 years",
                    ">18-29 years"))

table(MasterFrame$Category)
```

## Get basic statistics

Get some basic statistics, like number of DMD cases and controls by site, age distribution for each
group:

```{r}
##number of DMD cases and controls:
table(MasterFrame$Status)

##number of people at each site:
table(MasterFrame$Site)

##number of DMD cases and controls by site:
table(MasterFrame$Status, MasterFrame$Site)

##age distribution in each group:
quantile(MasterFrame$Age[MasterFrame$Status == "DMD"])
quantile(MasterFrame$Age[MasterFrame$Status == "Control"])

table(MasterFrame$Category[MasterFrame$Status == "DMD"])
table(MasterFrame$Category[MasterFrame$Status == "Control"])
```

## Generate Table 1

Generate Table 1 from paper:
```{r}
tabBySite <- table(MasterFrame$Site, MasterFrame$Status)
tabBySite <- data.frame(tabBySite)
tabBySite
colnames(tabBySite) <- c("Site", "Status", "N")
reshape(tabBySite, timevar="Status", idvar="Site", direction="wide")
tabBySite <- reshape(tabBySite, timevar="Status", idvar="Site", direction="wide") 
##reorder to put DMD first
tabBySite
tabBySite <- tabBySite[,c(1,3,2)]
##reorder to sort by total numbers
tabBySite <- tabBySite[order(rowSums(as.matrix(tabBySite[,-1])),
                             decreasing=TRUE), ]

##now have a separate table for age
tabByAge <- 
  data.frame(Age = "Age",
             DMD = 
               paste(round(quantile(MasterFrame$Age[MasterFrame$Status == "DMD"],
                                    probs = c(0, 0.5, 1)),1), sep="", collapse=", "),
             Controls = 
               paste(round(quantile(MasterFrame$Age[MasterFrame$Status == "Control"],
                                    probs = c(0, 0.5, 1)),1), sep="", collapse=", "))

tabByAge                                  
##add in total by age categories
tabByAgeGroup <- table(MasterFrame$Category, MasterFrame$Status)
tabByAgeGroup <- data.frame(tabByAgeGroup)
tabByAgeGroup
colnames(tabByAgeGroup) <- c("Age", "Status", "N")
reshape(tabByAgeGroup, timevar="Status", idvar="Age", direction="wide")
tabByAgeGroup <- reshape(tabByAgeGroup, timevar="Status", idvar="Age", direction="wide") 
##reorder to put DMD first
tabByAgeGroup
tabByAgeGroup <- tabByAgeGroup[,c(1,3,2)]
colnames(tabByAge) <- c("Age", "N.DMD", "N.Control")
tabByAge <- sapply(tabByAge, as.character)
tabByAgeGroup <- sapply(tabByAgeGroup, as.character)
tabByAge <- rbind(tabByAge, tabByAgeGroup)
tabByAge

##add in totals at the end
tabByAge <- 
  rbind(tabByAge,
        c("Total", sum(MasterFrame$Status == "DMD"), 
          sum(MasterFrame$Status == "Control")))
tabByAge

rownames(tabBySite) <- rownames(tabByAge) <- NULL
```

```{r}
kable(tabBySite)
kable(tabByAge)
```

## Save objects

```{r}
save(list=c("MasterFrame","mets"),
     file="../objs/MasterFrame.RData")
```

## Session information

Get session information:

```{r}
sessionInfo()
```